---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, echo=TRUE}
library("plyr")
library("dplyr")
library("lubridate")
library("lattice")
```

### Read activity data file
```{r, echo=TRUE}
where<-getwd()
print(where)
activitydata<-read.csv("activity.csv")
adstr<-str(activitydata)
adhead<-head(activitydata)
print(adstr)
print(adhead)
```

## What is mean total number of steps taken per day?
### sum number steps per day

```{r, echo=TRUE}
StepsbyDate <-aggregate(activitydata$steps ~ activitydata$date, data=activitydata, FUN=sum)
colnames(StepsbyDate)<-c("date","steps")
maxSBD<-range(StepsbyDate$steps)
maxval<-maxSBD[2]+5000
breaks= seq(5000,maxval, by=5000)
```
## plot histogram based on steps per day

```{r, echo=TRUE}

SBD<-StepsbyDate$steps
SBD.cut = cut(SBD,breaks,right=FALSE)
SBD.cut
SBD.freq=table(SBD.cut)
hist(SBD,main="Total Number Steps Taken per Day")
SBDmean<-mean(SBD)
SBDmedian<-median(SBD)
print(SBDmean)
print(SBDmedian)
```

## What is the average daily activity pattern?

```{r,echo=TRUE}
StepsbyInterval <-aggregate(activitydata$steps ~ activitydata$interval, data=activitydata,  FUN=ave)
colnames(StepsbyInterval)<-c("interval","steps")
SBI<-data.frame()
SBI<-cbind(StepsbyInterval$interval,StepsbyInterval$steps[,1])
plot(x=SBI[,1],y=SBI[,2],type="l", main="Average Daily Steps/Activity",xlab="Interval",ylab="Average number of steps")
```

### determine interval with max steps
```{r,echo=TRUE}
MaxSteps<-max(SBI[,2])
MaxInterval<-SBI[which.max(SBI[,2])]
print(MaxSteps)
print(MaxInterval)
```
## Imputing missing values
### determine NA values for Steps, data and interval 

```{r,echo=TRUE}
sum(is.na(activitydata$steps))
sum(is.na(activitydata$date))
sum(is.na(activitydata$interval))
total<-sum(is.na(activitydata$steps))+sum(is.na(activitydata$date))+sum(is.na(activitydata$interval))
```
### Total missing values =
```{r,echo=TRUE}
print(total)
```
### create copy of the dataset and fill in NA values with the overall data mean determined above.

```{r}
newval<-0
activitydata2<-activitydata
allobs<-length(activitydata2$steps)
DailyAvg <- activitydata2 %>% filter(steps >= 0) %>% 
    group_by(date) %>% summarize(total = sum(steps), avg = mean(steps))
for (i in 1:allobs) {
  if (is.na(activitydata2$steps[i])){
  activitydata2$steps[i]<-DailyAvg[,2]
  } else {
    }
  }

```

### recreate the histogram with the new dataset
```{r, echo=TRUE}
StepsbyDate <-aggregate(activitydata$steps ~ activitydata$date, data=activitydata, FUN=sum)
colnames(StepsbyDate)<-c("date","steps")
maxSBD<-range(StepsbyDate$steps)
maxval<-maxSBD[2]+5000
breaks= seq(5000,maxval, by=5000)
SBD2<-StepsbyDate$steps
SBD2.cut = cut(SBD,breaks,right=FALSE)
SBD2.freq=table(SBD.cut)
hist(SBD,main="Dataset with Mean Replacement for NA - Total Number Steps Taken per Day")
SBDmean<-mean(SBD)
SBDmedian<-median(SBD)
print(SBDmean)
print(SBDmedian)
```
## Are there differences in activity patterns between weekdays and weekends?

### separate weekdays and weekend days.

```{r, echo= TRUE}
activitydata$DayoWeek<-weekdays(as.Date(activitydata$date))

activitydata$DayType<-with(activitydata,ifelse(DayoWeek == "Saturday" | DayoWeek == "Saturday", "weekend", "weekday"))
```
## plot the steps by time series weekend vs weekday

``` {r, echo=TRUE}
group_by(activitydata,DayType,interval)
xyplot(activitydata$steps ~ activitydata$interval | activitydata$DayType, xlab = "Interval", ylab="Average # of Steps",type="l", layout=c(1,2))
```